Over-relaxed Sinkhorn's Algorithm for Regularized Optimal Transport

### 2017-10-06 paper.tex
Add proof of global convergence

\[
K^\epsilon(\gamma) = \scal{c}{\gamma}
+ \epsilon \sum_{i,j} \gamma_{i,j} (\log(\gamma_{i,j}) - 1)
\]
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma^*)
\end{equation}
Constraint sets:
\begin{align*}
A_1 &: \begin{cases}
\IR^{mn} \rightarrow \IR^m \\
(A_1 x)_i = \sum_j x_{i,j}
\end{cases} &
A_2 &: \begin{cases}
\IR^{mn} \rightarrow \IR^n\\
(A_2 x)_j = \sum_i x_{i,j}
\end{cases}
\end{align*}
\begin{align*}
\Ccal_1 &= \left\{ \gamma\in\IR^{mn} \mid A_1 \gamma = \mu \right\} &
\Ccal_2 &= \left\{ \gamma\in\IR^{mn} \mid A_2 \gamma = \nu \right\}
\end{align*}
Kullback-Leibler divergence:
\[ 
KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}
\]
Bregman projection onto $\Ccal$:
\[
P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} KL(\gamma,\xi)
\]
\[\argmin_{\gamma \in \Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma) = P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon})\]
Optimal transport plan: Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ onto $\Ccal_1 \cap \Ccal_2$.

@@

\begin{align*}
\gamma^0 &= e^{-c/\epsilon} &
\gamma^{k+1} = P_{\Ccal_2}(P_{\Ccal_1}(\gamma^k))
\end{align*}
Bregman's result: iterative Bregman projections onto $\Ccal_1$ and $\Ccal_2$ converge to a point in the intersection $\Ccal_1 \cap \Ccal_2$ \cite{bregman67}.

Hence this algorithm is correct:
\[\lim_{k\rightarrow +\infty} \gamma^k = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0) = \gamma^*\]

@@

Projections actually correspond to diagonal scalings.
\begin{align*}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &
a &=  \frac{\mu}{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &
b &= \frac{\nu}{A_2 \gamma}
\end{align*}
(pointwise division).

In practice, simply store $(a^k, b^k)$:
\[
\gamma^k = \diag(a^k) \gamma^0 \diag{b^k}
\]
\begin{align*}
a^{k+1} &= \frac{\mu}{\gamma^0 b^k} &
b^{k+1} &= \frac{\nu}{^t \gamma^0 a^{k+1}} 
\end{align*}

@@

\subsection{Lyapunov function}
Let $\gamma^*$ denote the solution of the regularized OT problem.
The function $F$ is defined as:
\begin{equation}\label{eq:lyapunov_function}
F(\gamma) = KL(\gamma^*, \gamma)
\end{equation}
We shall use it as a Lyapunov function, so that $F(\gamma^{k+1}) < F(\gamma^k)$ as long as the process has not converged.

Note that it is not necessary to know $\gamma^*$ in order to calculate the difference $F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma))$, as shown by the following lemma.
\begin{lemma}\label{lemma:lyapunov_decrease}
	Take $\gamma$ in $\IR^{mn}_{+*}$. The decrease in value of the Lyapunov function can be calculated with the following formula:
	\begin{equation} \label{eq:kl_diff_scal}
	F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma)) = 
	\scal{\mu}{\varphi_\omega \left(\frac{A_1 \gamma}{\mu}\right)},
	\end{equation}
	where
	\begin{equation}
	\varphi_\omega(x) = x(1-x^{-\omega}) - \omega e^x
	\end{equation}
	is a real function, applied coordinate-wise.
\end{lemma}
Thus, calculating the decrease in value of the function $F$ is relatively cheap and straightforward.

@@

\begin{theorem}
	Let $\theta$ and $\omega$ be such that $1\le \theta < \omega < 2$. Let $U$ be the open set defined by:
	\begin{equation}\label{eq:open_set_U}
	U = \left\{
	\gamma \in \IR_{+*}^{nm} \mid
	F(P^\omega_{\Ccal_1}(\gamma)) < F(\gamma)
	\text{ and }
	F(P^\omega_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma))) < F(P^\theta_{\Ccal_1}(\gamma))
	\right\}.
	\end{equation}
	Set the initial value $\gamma^0 = e^{-c/\epsilon}$, and iterate:
	\begin{align*}
	\gamma^{k+1} =
	\begin{cases}
	P^\theta_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma^k)) & \text{if } \gamma \in U \\
	P_{\Ccal_2}(P_{\Ccal_1})(\gamma^k) & \text{otherwise.}
	\end{cases}
	\end{align*}
	Then the iterates $(\gamma^k)$ converge to $\gamma^*$.
\end{theorem}
{\color{red} $\longrightarrow$ Has to calculate 3 projections sometimes. This result is weaker than the one from my report, but easier to prove. Maybe we could avoid having $U$, and rather perform the test independently on each variable ? (Or we could find a cheaper test).}
\begin{remark}
	Given lemma \ref{lemma:lyapunov_decrease}, we can see that $\gamma^* \in U$. Therefore, iterations eventually all use the over-relaxed operators.
\end{remark}

\begin{proof}
	(New proof, using the Lyapunov function. Much shorter and cleaner than the one from my report.)
\end{proof}

>>>>

\[
K^\epsilon(\gamma) = \scal{c}{\gamma}
+ \epsilon \sum_{i,j} \gamma_{i,j} (\log(\gamma_{i,j}) - 1)
\]
Two discrete probability measures $\mu^k \in \IR_{+*}^{n_k}$.
\begin{equation}\label{eq:reg_ot_pb}
\gamma^* = \argmin_{\Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma^*)
\end{equation}
Constraint sets:
\begin{align*}
A_1 &: \begin{cases}
\IR^{n_1 n_2} \rightarrow \IR^{n_1} \\
(A_1 x)_i = \sum_j x_{i,j}
\end{cases} &
A_2 &: \begin{cases}
\IR^{n_1 n_2} \rightarrow \IR^{n_2}\\
(A_2 x)_j = \sum_i x_{i,j}
\end{cases}
\end{align*}
\begin{align*}
\Ccal_k &= \left\{ \gamma\in\IR^{n_1 n_2} \mid A_k \gamma = \mu^k \right\}
\end{align*}
Kullback-Leibler divergence:
\[ 
KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}
\]
Bregman projection onto $\Ccal$:
\[
P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} KL(\gamma,\xi)
\]
\[\argmin_{\gamma \in \Ccal_1 \cap \Ccal_2} K^\epsilon(\gamma) = P_{\Ccal_1 \cap \Ccal_2} (e^{-c/\epsilon})\]
Optimal transport plan: Bregman projection of $\gamma^0 = e^{-c/\epsilon}$ onto $\Ccal_1 \cap \Ccal_2$.

@@

\begin{align*}
\gamma^0 &= e^{-c/\epsilon} &
\gamma^{l+1} = P_{\Ccal_2}(P_{\Ccal_1}(\gamma^l))
\end{align*}
Bregman's result: iterative Bregman projections onto $\Ccal_1$ and $\Ccal_2$ converge to a point in the intersection $\Ccal_1 \cap \Ccal_2$ \cite{bregman67}.

Hence this algorithm is correct:
\[\lim_{l\rightarrow +\infty} \gamma^l = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0) = \gamma^*\]

@@

Projections actually correspond to diagonal scalings.
\begin{align*}
P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &
a &=  \frac{\mu^1}{A_1 \gamma} \\
P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &
b &= \frac{\mu^2}{A_2 \gamma}
\end{align*}
(pointwise division).

In practice, simply store $(a^l, b^l)$:
\[
\gamma^l = \diag(a^l) \gamma^0 \diag{b^l}
\]
\begin{align*}
a^{l+1} &= \frac{\mu^1}{\gamma^0 b^l} &
b^{l+1} &= \frac{\mu^2}{^t \gamma^0 a^{l+1}} 
\end{align*}

@@

\subsection{Lyapunov function}
Let $\gamma^*$ denote the solution of the regularized OT problem.
The function $F$ is defined as:
\begin{equation}\label{eq:lyapunov_function}
F(\gamma) = KL(\gamma^*, \gamma)
\end{equation}
It shall be used as a Lyapunov function, so that $F(\gamma^{k+1}) < F(\gamma^k)$ as long as the process has not converged.

\begin{lemma} \label{lemma:KL_compact}
	For any $M \in \IR_+^*$, the sublevel set $\left\{ \gamma \mid F(\gamma) \le M \right\}$ is compact.
\end{lemma}
\begin{proof}
	{\color{red} TODO}
\end{proof}

Note that it is not necessary to know $\gamma^*$ in order to calculate the difference $F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma))$, as shown by the following lemma.
\begin{lemma}\label{lemma:lyapunov_decrease}
	Take $\gamma$ in $\IR^{mn}_{+*}$. The decrease in value of the Lyapunov function can be calculated with the following formula:
	\begin{equation} \label{eq:kl_diff_scal}
	F(\gamma) - F(P^\omega_{\Ccal_k}(\gamma)) = 
	\scal{\mu}{\varphi_\omega \left(\frac{A_1 \gamma}{\mu}\right)},
	\end{equation}
	where
	\begin{equation}
	\varphi_\omega(x) = x(1-x^{-\omega}) - \omega \log x
	\end{equation}
	is a real function, applied coordinate-wise.
\end{lemma}
Thus, calculating the decrease in value of the function $F$ is relatively cheap and straightforward.

@@

\begin{theorem}
	Let $\theta$ and $\omega$ be such that $1\le \theta < \omega < 2$. Let $U$ be the open set defined by:
	\begin{equation}\label{eq:open_set_U}
	U = \left\{
	\gamma \in \IR_{+*}^{nm} \mid
	F(P^\omega_{\Ccal_1}(\gamma)) < F(\gamma)
	\text{ and }
	F(P^\omega_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma))) < F(P^\theta_{\Ccal_1}(\gamma))
	\right\}.
	\end{equation}
	Set the initial value $\gamma^0 = e^{-c/\epsilon}$, and iterate:
	\begin{align*}
	\gamma^{l+1} =
	\begin{cases}
	P^\theta_{\Ccal_2}(P^\theta_{\Ccal_1}(\gamma^l)) & \text{if } \gamma \in U \\
	P_{\Ccal_2}(P_{\Ccal_1})(\gamma^k) & \text{otherwise.}
	\end{cases}
	\end{align*}
	Then the iterates $(\gamma^l)$ converge to $\gamma^*$.
\end{theorem}
{\color{red} $\longrightarrow$ Has to calculate 3 projections sometimes. This result is weaker than the one from my report, but easier to prove. Maybe we could avoid having $U$, and rather perform the test independently on each variable ? (Or we could find a cheaper test).}
\begin{remark}
	Given lemma \ref{lemma:lyapunov_decrease}, it can be seen that $\gamma^* \in U$. Therefore, iterations eventually all use the over-relaxed operators.
\end{remark}

\begin{lemma}
	\label{lemma:trivial_intersection}
	Let us take $\gamma^0$ in $\IR_{+*}^{n_1 n_2}$,
	and denote
	\[
	S = \left\{
	\diag(a) \gamma^0 \diag(b),\quad
	(a,b) \in \IR_{+*}^{n_1 + n_2}
	\right\}
	\]
	the set of matrices that are diagonally similar to $\gamma^0$.
	Then the set $S \cap \Ccal_1 \cap \Ccal_2$ contains exactly one element $\gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0)$.
\end{lemma}

\begin{lemma}\label{lemma:F_P_theta}
	Let $1\le \theta < \omega$. Then, for any $\gamma \in \IR_{+*}^{nm}$, one has
	\begin{equation}\label{eq:F_P_theta}
	F(P^\theta_{\Ccal_k}(\gamma)) \le F(P^\omega_{\Ccal_k}(\gamma)).
	\end{equation}
	Moreover, equality occurs if and only if $\gamma \in \Ccal_k$.
\end{lemma}
\begin{proof}
	Thanks to lemma \ref{lemma:lyapunov_decrease}, one knows that
	\[
	F(P^\theta_{\Ccal_k}(\gamma)) - F(P^\omega_{\Ccal_k}(\gamma))
	= \scal{\mu}{(\varphi_\omega - \varphi_\theta) \left( \frac{A_k \gamma}{\mu} \right) } .
	\]
	Moreover, $t \in [1,\infty) \mapsto \varphi_t(x)$ is nonincreasing:
	\[
	\frac{d}{dt} \varphi_t(x) = \log x (x^{1-t} - 1).
	\]
	For $x\neq 1$, it is even strictly decreasing.
	Thus the inequality (\ref{eq:F_P_theta}) is valid, with equality \emph{iff} $A_1 \gamma = \mu$.
\end{proof}

\begin{proof}[Proof of the theorem]
	(New proof, using the Lyapunov function. Much shorter and cleaner than the one from my report.)
	First of all, notice that the operators $P_{\Ccal_k}^\theta$ apply a scaling to lines or columns of matrices. All $(\gamma^l)$ are thus diagonally similar to $\gamma^0$:
	\[
	\forall l\ge0,\quad \gamma^l \in S
	\]

	The set $U$ is constructed in such a way that, by applying lemma \ref{lemma:F_P_theta} several times, one ensures that the sequence $(F(\gamma^l))$ is nonincreasing. Moreover, it is strictly decreasing as long as $\gamma^l \not \in \Ccal_1 \cap \Ccal_2$.
	Hence, 
	\[
	F(\gamma^l) \le F(\gamma^0).
	\]
	By lemma \ref{lemma:KL_compact}, the sequence $(\gamma^l)$ is thus precompact. Let us show that any accumulation point $\zeta$ belongs to $\Ccal_1 \cap \Ccal_2$. Since $S$ is closed in $\IR_{+*}^{n_1 n_2}$, by lemma \ref{lemma:trivial_intersection} this would mean that $\zeta = \gamma^*$, and would be sufficient to prove the convergence.

	Let $(\gamma^{k_l})$ be a subsequence of $(\gamma^l)$ that converges to $\zeta$.
	Let us denote, for convenience, $T_1 = P_{\Ccal_2}^\theta \circ P_{\Ccal_1}^\theta$ and $T_2 = P_{\Ccal_2} \circ P_{\Ccal_1}$ the two operators used in the iteration, and $T$ the compound operator, so that $T_{|U} \equiv T_{1|U}$ and $T_{|U^c} \equiv T_{2|U^c}$.
	\begin{itemize}
		\item If $\zeta \in U$, then since $U$ is open, all $(\gamma^{k_l})$ are eventually in $U$ as well.
		Thus, for any large enough $l$, $ T(\gamma^{k_l}) = T_1(\gamma^{k_l}) $. By continuity of $T_1$ on $U$, since $(\gamma^{k_l})$ converges to $\zeta$, 
		\[T(\zeta) = \lim_{l\rightarrow \infty} T(\gamma^{k_l}).\] Thus
		 \[F(T(\zeta)) = \lim_{l \rightarrow \infty} F(T(\gamma^{k_l})) \ge \lim_{l\rightarrow \infty} F(\gamma^{k_{l+1}}) .\]
		 \[
		 F(T(\zeta)) \ge F(\zeta)
		 \] Since $\zeta$ is inside the set $U$, which ensures strict decrease of the Lyapunov function, this is absurd.
		 \item If $\zeta \in U^c$ and an infinite number of $(\gamma^{k_l})$ are in $U^c$, then the same argument of continuity applied to $T_2$ yields
		 \[
		 F(T(\zeta)) = F(\zeta).
		 \]
		 The function $\varphi_1$ from lemma \ref{lemma:lyapunov_decrease} is strictly positive except at $x=1$. Since the case of equality occurs for both projections, this ensures that $\zeta \in \Ccal_1 \cap \Ccal_2$.

		 \item If $\zeta \in U^c$ and all but finitely many $\gamma^{k_l}$ are in $U$, then $\zeta$ is on the border $\partial U$. In particular it belongs to $\bar{U}$, and therefore
		 \[
		 F(P^\omega_{\Ccal_1}(\zeta)) \le F(\zeta) \quad
		 \text{ and } \quad
		 F(P^\omega_{\Ccal_2}(P^\theta_{\Ccal_1}(\zeta))) \le F(P^\theta_{\Ccal_1}(\zeta)).
		 \]
		 By lemma \ref{lemma:F_P_theta}, the inequality \begin{equation} \label{eq:ineq1}
		 F(P_{\Ccal_1}^\theta(\zeta)) \le F(P_{\Ccal_1}^\omega(\zeta)) \end{equation}
		 is valid; it is strict unless $\zeta \in \Ccal_1$.
		 The same lemma yields
		 \begin{equation} \label{eq:ineq2}
		 F(P_{\Ccal_2}^\theta(P_{\Ccal_1}^\theta(\zeta)) \le F(P_{\Ccal_2}^\omega(P_{\Ccal_1}^\theta(\zeta)), \end{equation}
		 with a strict inequality unless $P_{\Ccal_1}^\theta(\zeta) \in \Ccal_2$.
		 To sum up, we have the series of inequalities:
		 \[
		 F(\zeta) 
		 \overset{\zeta \in \bar{U}}{\ge}
		 F(P^\omega_{\Ccal_1}(\zeta))
		 \overset{\text{(\ref{eq:ineq1})}}{\ge}
		 F(P^\theta_{\Ccal_1}(\zeta))
		 \overset{\zeta \in U}{\ge}
		 F(P_{\Ccal_2}^\omega(P_{\Ccal_1}^\theta(\zeta)))
		 \overset{\text{(\ref{eq:ineq2})}}{\ge}
		 F(P_{\Ccal_2}^\theta(P_{\Ccal_1}^\theta(\zeta)))
		 \]

		 Yet, by continuity of $T_1 = P_{\Ccal_2}^\theta \circ P_{\Ccal_1}^\theta $ on $\bar{U}$, we have
		 \[
		 F(T_1(\zeta)) = F(\zeta).
		 \]
		 Thus both (\ref{eq:ineq1}) and (\ref{eq:ineq2}) fall in the case of equality, which means $\zeta \in \Ccal_1$ and $P^\theta_{\Ccal_1}(\zeta) = \zeta \in \Ccal_2$.
	\end{itemize}
\end{proof}
